Hierarchical Time Series Forecasting in Emergency Medical Services

### 2023-07-19 manuscript/05-results.qmd
Minor rewording, typos and fixed some refs

In this section, we compare the forecasting performance of the Naive, ETS, GLM, and TSGLM models along with the ensemble, using base forecast and Minimum Trace (MinT) reconciliation methods. We have also computed the forecast accuracy for Ordinary Least Square (OLS) and Weighted Least Square (WLS) approaches, along with bottom-up forecasting. However, they are not reported here because their accuracy is outperformed by MinT. We should also note that forecasts, and consequently their corresponding errors, are generated for the entire hierarchy and they could be reported at any level, if required. But to save space, we have reported only the top level (Total), the bottom level, and the levels corresponding to Control areas and Health boards. The latters are chosen because this is where decision-making takes place, so these forecasts are the most important.

The overall forecasting performance is reported in @tbl-result, in which the average forecast accuracy over horizons 43--84 days (corresponding to the planning horizon) is presented per model, method, and the hierarchical level. Reported forecast accuracy is averaged across all forecast horizons, rolling origins, and series at each level. @tbl-result presents both point and probabilistic forecast accuracy at total, control area, health board and bottom level series. Point forecast performance is reported using MASE and MSSE in @tbl-result-1  and [-@tbl-result-2], respectively. Probabilistic forecast accuracy is reported using CRPS in @tbl-result-3. The bold entries in each table identify a combination of method and model that performs best for the corresponding level (i.e. each column), based on the smallest values of accuracy measures.

@tbl-result-1 and [-@tbl-result-2] show that forecast reconciliation (i.e. MinT) improves forecast accuracy at the higher levels of the hierarchy including total, control area and health board. However, it does not result in accuracy improvement at the bottom level series, for which base forecasts are more accurate. This might be due to the noisy structure of time series at the bottom level, and possibly due to very different patterns in the aggregated series. It is also clear from @tbl-result-1 that the ensemble method improves forecast accuracy at total, control area and health board. However, this does not remain valid for bottom series where different individual methods perform best, depending on the accuracy measure. \textcolor{teal}{While the forecast reconciliation approach aims to enhance forecast accuracy, its effectiveness may not always be guaranteed, especially if the bottom-level series exhibit excessive noise and lack systematic patterns. Despite this, reconciling forecasts at the bottom level can offer advantages by generating coherent forecasts that facilitate alignment in planning across various teams within an organization, promote better coordination, and prevent conflicting decisions. Moreover, even when dealing with noisy and irregular bottom-level series, reconciliation can still improve forecast accuracy at higher levels of the hierarchy by leveraging the information available across the hierarchy. Therefore, although the bottom-level forecasts may not be highly accurate on their own, reconciling them with higher-level forecasts can still provide a more consistent view of future demand and potentially yield more accurate forecasts at other levels.}

@@

msse1 <- rmsse |>
  filter(
    method %in% c("mint", "base"),
    model != "qcomb",
    h > 42
  ) |>
  mutate(model = factor(model,
    levels = c("naiveecdf","ets","iglm","tscount","ensemble","ensemble2"),
    labels = c("Naive", "ETS", "GLM", "TSGLM","Ensemble","Ensemble2")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT")),
  ) |>
  group_by(method, model, series) |>
  summarise(msse = mean(msse), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = msse) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
mase1 <- mase |>
  filter(
    method %in% c("mint", "base"),
    model != "qcomb",
    h > 42
  ) |>
  mutate(
    model = factor(model,
      levels = c("naiveecdf","ets","iglm","tscount","ensemble","ensemble2"),
      labels = c("Naive", "ETS", "GLM", "TSGLM","Ensemble","Ensemble2")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
  ) |>
  group_by(method, model, series) |>
  summarise(mase = mean(mase), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = mase) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
crps1 <- readr::read_rds(here::here("results/crps.rds")) |>
  filter(method %in% c("mint", "base"),
         model != "qcomb",
         h > 42
  ) |>
  mutate(model = factor(model,
    levels = c("naiveecdf","ets","iglm","tscount","ensemble","ensemble2"),
    labels = c("Naive", "ETS", "GLM", "TSGLM","Ensemble","Ensemble2")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
  ) |>
  group_by(method, model, series) |>
  summarise(crps = mean(crps), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = crps) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()

@@

```{r}
#| label: fig-accuracy
#| fig-width: 8
#| fig-height: 5
#| out.width: "100%"
#| fig-cap: "Average accuracy by week for 12 weeks using MinT reconciliation. The total number of days used to calculate the accuracy in the test set is 462. Forecasts are generateled every 42 days, therefore we use 11 samples to calculate the average accuracy. CRPS is relative to a naive Empirical Cumulative Distribution Function (ECDF). MASE and MSSE are relative to the corresponding values for the training set."
#| fig-pos: "H"
crps <- crps |>
  pivot_longer(ensemble:tscount, names_to = "model", values_to = "crps")

@@

# Plot of average accuracy vs week for each method for Total
acc_summary <- accuracy |>
  filter(
    series %in% c("Total", "Control areas", "Health boards"),
    method == "mint", model != "qcomb"
  ) |>
  mutate(model = case_when(
    model == "naiveecdf" ~ "Naive",
    model == "ets" ~  "ETS",
    model == "tscount" ~ "TSGLM",
    model == "iglm" ~ "GLM",
    model == "ensemble" ~ "Ensemble",
    model == "ensemble2" ~ "Ensemble2"
  )) |> 
  mutate(
    measure = factor(measure, levels = c("mase","msse","crps"), labels = c("MASE","MSSE","CRPS")),
    series = factor(series, levels = c("Total", "Control areas", "Health boards")),
    model = factor(model, levels = c("Naive", "ETS", "TSGLM", "GLM", "Ensemble", "Ensemble2")),
    week = factor(trunc((h - 1) / 7) + 1)
  ) |>
  group_by(week, model, measure, series) |>
  summarise(accuracy = mean(accuracy), .groups = "drop")

@@

In addition to the overall forecast accuracy presented in @tbl-result, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-accuracy. The figure focuses on the hierarchical levels important for decision-making including total, control area, and health board; however, the accuracy could be calculated for any level. We only illustrate the results of the MinT method, given its strong performance described in @tbl-result. For illustration purposes, we report the average weekly forecast accuracy instead of the daily forecast horizon, as this reduces the visual noise in the figure. Thus, the x-axis shows horizons from week 1 ($h= 1,\dots,7$) to week 12 ($h= 78,\dots,84$). The forecast horizon from week 7 to week 12 corresponds to the upcoming planning horizon, which is used by planners and decision-makers. For both the point forecast and distributional accuracy  we can see that the ensemble approach performs best across almost all horizons, with the biggest differences at the highest levels of aggregation. It is important to highlight that, all forecasting models outperform the Naive empirical distribution that is used as a benchmark for both point and probabilistic forecasts.

\textcolor{teal}{Despite using Poisson regression models to create count distributions of attended incidents for the base forecasts, it is important to note that the reconciled forecast distributions do not maintain a count format. In practical scenarios, there might be a need to use integer forecasts. Count forecast reconciliation is an active area of research, and it would be interesting to explore how our approach could be adapted to generate count-reconciled probabilistic forecasts in future studies. One possible approach to address this is by rounding the forecasts. However, the impact of rounding on forecast accuracy varies depending on the level of hierarchy and the scale of the data. In situations with high-volume demand, the effects of rounding may become negligible, and forecast accuracy calculations can overlook integer effects. On the other hand, in low-volume demand settings, such as forecasts at the bottom level of the hierarchy, integer (rounding) effects may have a more noticeable influence on forecast accuracy.}

@@

@fig-forecast-density \textcolor{teal}{provides an illustrative example of a probabilistic forecast for future demand, one day ahead, regarding the total attended incidents in the PO health board. The black dashed line in the graph represents the point forecast, which presents a single numerical estimate for the future number of attended incidents. Due to the complexity of including such plots for the entire hierarchy and 84 days ahead, only one example is presented here for 1 day ahead. However, it is feasible to generate these plots for the entire hierarchy and for any forecast horizon if necessary.}

@@

\textcolor{teal}{The purpose of probabilistic forecasting, as demonstrated in} @fig-forecast-density, \textcolor{teal}{is to quantify uncertainty. Decisions based on these forecasts could focus on the tails of the distribution: unexpected high demand leading to crowding and inefficiency, or unexpectedly low demand resulting in wasted resources. Such forecasts are valuable tools for decision-makers and planners, especially when dealing with low-probability, high-cost situations. Different EMS managements may have varying risk attitudes depending on resource availability, making it crucial to consider the entire distribution when making decisions. For instance, these forecasts enable management to calculate the probability of demand exceeding a certain threshold of available resources (e.g., 90\%), which can serve as an informative early warning measure for crowdedness.}

@@

```{r}
#| label: fig-forecast-density
#| fig-width: 8
#| fig-height: 5
#| out.width: "100%"
#| fig-cap: "An illustrative example of forecast distribution of ambulance demand (i.e. total incidence attended) in PO health board for 1 day ahead. x-axis shows all possible outcomes that may occur, with thier likelihhod in y-aixs"
#| fig-pos: "H"
incident_gthf <- readr::read_rds(here::here("data/incidents_gt.rds"))

>>>>

In this section, we compare the forecasting performance of the Stationary, ETS, GLM, and TSGLM models along with the ensemble, using base forecast and Minimum Trace (MinT) reconciliation methods. We have also computed the forecast accuracy for Ordinary Least Square (OLS) and Weighted Least Square (WLS) approaches, along with bottom-up forecasting. However, they are not reported here because their accuracy is outperformed by MinT. We should also note that forecasts, and consequently their corresponding errors, are generated for the entire hierarchy and they could be reported at any level, if required. But to save space, we have reported only the top level (Total), the bottom level, and the levels corresponding to Control areas and Health boards. The latters are chosen because this is where decision-making takes place, so these forecasts are the most important.

The overall forecasting performance is reported in @tbl-result, in which the average forecast accuracy over horizons 43--84 days (corresponding to the planning horizon) is presented per model, method, and the hierarchical level. Reported forecast accuracy is averaged across all forecast horizons, rolling origins, and series at each level. @tbl-result presents both point and probabilistic forecast accuracy at total, control area, health board and bottom-level series. Point forecast performance is reported using MASE and MSSE in @tbl-result-1  and [-@tbl-result-2], respectively. Probabilistic forecast accuracy is reported using CRPS in @tbl-result-3. The bold entries in each table identify a combination of method and model that performs best for the corresponding level (i.e. each column), based on the smallest values of accuracy measures.

@tbl-result-1 and [-@tbl-result-2] show that forecast reconciliation (i.e. MinT) improves forecast accuracy at the higher levels of the hierarchy including total, control area and health board. However, it does not result in accuracy improvement at the bottom-level series, for which base forecasts are more accurate. This might be due to the noisy structure of time series at the bottom level, and possibly due to very different patterns in the aggregated series. It is also clear from @tbl-result-1 that the ensemble method improves forecast accuracy at total, control area and health board. However, this does not remain valid for bottom series where different individual methods perform best, depending on the accuracy measure. \textcolor{teal}{While the forecast reconciliation approach aims to enhance forecast accuracy, its effectiveness is not guaranteed, especially if the bottom-level series exhibit excessive noise and lack systematic patterns. Despite this, reconciling forecasts at the bottom level can offer advantages by generating coherent forecasts that facilitate alignment in planning across various teams within an organization, promote better coordination, and prevent conflicting decisions. Moreover, even when dealing with noisy and irregular bottom-level series, reconciliation can still improve forecast accuracy at higher levels of the hierarchy by leveraging the information available across the hierarchy. Therefore, although the bottom-level forecasts may not be highly accurate on their own, reconciling them with higher-level forecasts can still provide a more consistent view of future demand and potentially yield more accurate forecasts at other levels.}

@@

msse1 <- rmsse |>
  filter(
    method %in% c("mint", "base"),
    model != "qcomb",
    h > 42,
    model != "ensemble2"
  ) |>
  mutate(model = factor(model,
    levels = c("naiveecdf","ets","iglm","tscount","ensemble","ensemble2"),
    labels = c("Stationary", "ETS", "GLM", "TSGLM","Ensemble","Ensemble2")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT")),
  ) |>
  group_by(method, model, series) |>
  summarise(msse = mean(msse), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = msse) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
mase1 <- mase |>
  filter(
    method %in% c("mint", "base"),
    model != "qcomb",
    h > 42,
    model != "ensemble2"
  ) |>
  mutate(
    model = factor(model,
      levels = c("naiveecdf","ets","iglm","tscount","ensemble","ensemble2"),
      labels = c("Stationary", "ETS", "GLM", "TSGLM","Ensemble","Ensemble2")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
  ) |>
  group_by(method, model, series) |>
  summarise(mase = mean(mase), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = mase) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()
crps1 <- readr::read_rds(here::here("results/crps.rds")) |>
  filter(method %in% c("mint", "base"),
         model != "qcomb",
         h > 42,
         model != "ensemble2"
  ) |>
  mutate(model = factor(model,
    levels = c("naiveecdf","ets","iglm","tscount","ensemble","ensemble2"),
    labels = c("Stationary", "ETS", "GLM", "TSGLM","Ensemble","Ensemble2")),
    method = factor(method, levels = c("base", "mint"), labels = c("Base", "MinT"))
  ) |>
  group_by(method, model, series) |>
  summarise(crps = mean(crps), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = crps) |>
  select(Method = method, Model = model, Total, `Control areas`, `Health boards`, Bottom) |>
  set_to_bold()

@@

```{r}
#| label: fig-accuracy
#| fig-width: 8
#| fig-height: 5
#| out.width: "100%"
#| fig-cap: "Average accuracy by week for 12 weeks using MinT reconciliation. The total number of days used to calculate the accuracy in the test set is 462. Forecasts are generateled every 42 days, therefore we use 11 samples to calculate the average accuracy. CRPS is relative to a stationary Empirical Cumulative Distribution Function (ECDF). MASE and MSSE are relative to the corresponding values for the training set."
#| fig-pos: "H"
crps <- crps |>
  pivot_longer(ensemble:tscount, names_to = "model", values_to = "crps")

@@

# Plot of average accuracy vs week for each method for Total
acc_summary <- accuracy |>
  filter(
    series %in% c("Total", "Control areas", "Health boards"),
    method == "mint", model != "qcomb"
  ) |>
  mutate(model = case_when(
    model == "naiveecdf" ~ "Stationary",
    model == "ets" ~  "ETS",
    model == "tscount" ~ "TSGLM",
    model == "iglm" ~ "GLM",
    model == "ensemble" ~ "Ensemble",
    model == "ensemble2" ~ "Ensemble2"
  )) |>
  mutate(
    measure = factor(measure, levels = c("mase","msse","crps"), labels = c("MASE","MSSE","CRPS")),
    series = factor(series, levels = c("Total", "Control areas", "Health boards")),
    model = factor(model, levels = c("Stationary", "ETS", "TSGLM", "GLM", "Ensemble", "Ensemble2")),
    week = factor(trunc((h - 1) / 7) + 1)
  ) |>
  group_by(week, model, measure, series) |>
  summarise(accuracy = mean(accuracy), .groups = "drop")

@@

In addition to the overall forecast accuracy presented in @tbl-result, we also report the point and probabilistic forecast accuracy measures for each forecast horizon in @fig-accuracy. The figure focuses on the hierarchical levels important for decision-making including total, control area, and health board; however, the accuracy could be calculated for any level. We only illustrate the results of the MinT method, given its strong performance described in @tbl-result. For illustration purposes, we report the average weekly forecast accuracy instead of the daily forecast horizon, as this reduces the visual noise in the figure. Thus, the x-axis shows horizons from week 1 ($h= 1,\dots,7$) to week 12 ($h= 78,\dots,84$). The forecast horizon from week 7 to week 12 corresponds to the upcoming planning horizon, which is used by planners and decision-makers. For both the point forecast and distributional accuracy  we can see that the ensemble approach performs best across almost all horizons, with the biggest differences at the highest levels of aggregation. It is important to highlight that, all forecasting models outperform the stationary empirical distribution that is used as a benchmark for both point and probabilistic forecasts.

\textcolor{teal}{Despite using Poisson regression models to create count distributions of attended incidents for the base forecasts, it is important to note that the reconciled forecast distributions do not maintain a count format. In practical scenarios, there might be a need to use integer forecasts. Count forecast reconciliation is an active area of research, and it would be interesting to explore how our approach could be adapted to generate count-reconciled probabilistic forecasts in future studies. One possible approach to address this is by rounding the forecasts. However, the impact of rounding on forecast accuracy varies depending on the level of hierarchy and the scale of the data. In situations with high-volume demand, the effects of rounding may be negligible, and forecast accuracy calculations can overlook integer effects. On the other hand, in low-volume demand settings, such as forecasts at the bottom level of the hierarchy, integer (rounding) effects may have a more noticeable influence on forecast accuracy.}

@@

@fig-forecast-density \textcolor{teal}{provides an illustrative example of a probabilistic forecast for future demand, one day ahead, of the total attended incidents in the PO health board. The black dashed line in the graph represents the point forecast (the mean of the distribution), which presents a single numerical estimate for the future number of attended incidents. Due to the complexity of including such plots for the entire hierarchy and 84 days ahead, only one example is presented here for 1 day ahead. However, it is feasible to generate these plots for the entire hierarchy and for any forecast horizon if necessary.}

@@

\textcolor{teal}{The purpose of probabilistic forecasting, as demonstrated in} @fig-forecast-density, \textcolor{teal}{is to quantify uncertainty. Decisions based on these forecasts could focus on the tails of the distribution: unexpectedly high demand leading to crowding and inefficiency, or unexpectedly low demand resulting in wasted resources. Such forecasts are valuable tools for decision-makers and planners, especially when dealing with low-probability, high-cost situations. Different EMS managements may have varying risk attitudes depending on resource availability, making it crucial to consider the entire distribution when making decisions. For instance, these forecasts enable management to calculate the probability of demand exceeding a certain threshold of available resources (e.g., 90\%), which can serve as an informative early warning measure for crowdedness.}

@@

```{r}
#| label: fig-forecast-density
#| fig-width: 8
#| fig-height: 5
#| out.width: "100%"
#| fig-cap: "An illustrative example of the forecast distribution of ambulance demand (i.e. total incidence attended) for the PO health board for 1 day ahead. The horizontal axis shows all possible outcomes that may occur, with their likelihood shown on the vertical axis. The dashed line indicates the mean of the distribution."
#| fig-pos: "H"
incident_gthf <- readr::read_rds(here::here("data/incidents_gt.rds"))

### 2023-07-19 manuscript/06-conclusion.qmd
Minor rewording, typos and fixed some refs

\textcolor{teal}{Although our study primarily focuses on Emergency Medical Services, it is essential to emphasize that the framework we propose has broad applicability across various service industries} [@ostrom2010moving]. \textcolor{teal}{Our approach is particularly valuable in situations where time series data is structured hierarchically and/or grouped, a common characteristic found in many sectors. This occurs when data can be naturally organized into different levels of hierarchies or when dependencies and relationships exist among entities within the system. For instance, in the supply chains } [@shugan2000advance], \textcolor{teal}{demand forecasting at different levels of the distribution network, such as regional warehouses or retail stores, is vital for efficient inventory management and minimizing stockouts. Our framework allows the reconciliation of forecasts, ensuring consistency and alignment throughout the supply chain, leading to improved decision-making and operational efficiency. In the financial industry} [@kimes1998strategic], \textcolor{teal}{where investments span multiple asset classes, geographical regions, or customer segments, our framework can be applied to forecast portfolio performance, asset allocation, or customer demand. Similarly, in transportation, the framework supports forecasting transportation demand at various levels, optimizing route planning and resource allocation. Likewise, in the hospitality and tourism industry} [@dekimpe2016impact], \textcolor{teal}{it facilitates forecasting demand rates at state, regional, and department levels, enabling strategic pricing, capacity planning, and revenue management for hotels and other travel-related businesses. Additionally, in call centers, accurate call volume forecasting at different levels of the call center hierarchy or grouped structure is crucial for workforce management and resource allocation. Implementing our framework, call centers can generate accurate forecasts for different skill groups, shifts, and locations, ensuring efficient staffing and optimal service levels to meet customer demands.}

>>>>

\textcolor{teal}{Although our study primarily focuses on Emergency Medical Services, it is essential to emphasize that the framework we propose has broad applicability across various service industries} [@ostrom2010moving]. \textcolor{teal}{Our approach is particularly valuable in situations where time series data is structured hierarchically and/or grouped, a common characteristic found in many sectors. This occurs when data can be naturally organized into different levels of hierarchies or when dependencies and relationships exist among entities within the system. For instance, in supply chains} [@shugan2000advance], \textcolor{teal}{demand forecasting at different levels of the distribution network, such as regional warehouses or retail stores, is vital for efficient inventory management and minimizing stockouts. Our framework allows the reconciliation of forecasts, ensuring consistency and alignment throughout the supply chain, leading to improved decision-making and operational efficiency. In the financial industry} [@kimes1998strategic], \textcolor{teal}{where investments span multiple asset classes, geographical regions, or customer segments, our framework can be applied to forecast portfolio performance, asset allocation, or customer demand. Similarly, in transportation, the framework supports forecasting transportation demand at various levels, optimizing route planning and resource allocation. Likewise, in the hospitality and tourism industry} [@dekimpe2016impact], \textcolor{teal}{it facilitates forecasting demand rates at state, regional, and department levels, enabling strategic pricing, capacity planning, and revenue management for hotels and other travel-related businesses. Additionally, in call centers, accurate call volume forecasting at different levels of the call center hierarchy or grouped structure is crucial for workforce management and resource allocation. Implementing our framework, call centers can generate accurate forecasts for different skill groups, shifts, and locations, ensuring efficient staffing and optimal service levels to meet customer demands.}
